Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
Changed the font before 4.1.

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether \lstinline{n'} is empty or contains the key \lstinline{k}.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p}, \texttt{r}, \texttt{n}.\ \texttt{pn} \mapsto (\texttt{p, n}) \ast \ \infp (\texttt{r}) \ast \infp (\texttt{n})
\end{mathpar}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(\texttt{r})$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

With this invariant in hand, the proof of \texttt{traverse} proceeds as follows. We begin by using the $\infp$ predicate to acquire the lock on \lstinline{pn->n} (line 4 in Figure \ref{traverse_giveup_b}). We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}. If the check fails (lines 18-20), we immediately release the lock and start over from the root node \lstinline{r}. Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}. In this case, we once again hold no locks, and have only the $\infp$ assertion indicating that the new node is in the data structure. % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \texttt{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

>>>>

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, \mathsf{v'}), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
\end{mathpar}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

With this invariant in hand, the proof of \texttt{traverse} proceeds as follows. We begin by using the $\infp$ predicate to acquire the lock on \lstinline{pn->n} (line 4 in Figure \ref{traverse_giveup_b}). We then call \lstinline{inRange} to check whether we are still on the path to a node that can hold \lstinline{k}. If the check fails (lines 18-20), we immediately release the lock and start over from the root node $r$. Otherwise, we proceed as in the lock-coupling template: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \texttt{k} belongs (and can satisfy the postcondition); if \texttt{findNext} returns 0, we have found the node containing \texttt{k} (and can satisfy the postcondition); otherwise, we release the lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \texttt{findNext}. In this case, we once again hold no locks, and have only the $\infp$ assertion indicating that the new node is in the data structure. % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \texttt{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.
